Separating random and deterministic sources of computational noise in explore-exploit decisions

### 2024-05-14 manuscript/2024 - Plot Comp Bio/manuscript_DeterministicRandomNoise.tex
added stats for fig 2CD
parameter recovery

	\subsubsection*{Model-based analysis}	
	Another approach to understanding behavior in the Horizon Task is to use a computational model. In this case, we model  participants' choices on the first free-choice trial by assuming they make decisions by computing the difference in value (or utility) $\Delta Q$ between the right and left options, choosing right when $\Delta Q > 0$ and left otherwise.  Specifically, we write
	\begin{equation}
		\label{eq:origmodel}
		\Delta Q= \Delta R+A \Delta I+b+n
	\end{equation}
	where, the experimentally controlled variables are $\Delta R=R_{right}-R_{left}$, the difference between the mean of rewards shown on the forced-choice trials, and $\Delta I$, the difference in information available for playing the two options on the first free-choice trial. For simplicity, and because information is manipulated categorically in the Horizon Task, we define $\Delta I$ to be +1 if one reward is drawn from the right option and three are drawn from the left in the [1 3] condition, -1 if one from the left and three from the right in the [1 3] condition, and in [2 2] condition, $\Delta I$ is 0.

@@

	The free parameters of this model are: the information bonus $A$, which controls the level of directed exploration; the noise standard deviation, $\sigma$, which controls the level of random exploration, and the spatial bias, $b$, which determines the extent to which participants prefer the option on the right. These free parameters are fit separately for each participant in each horizon and information condition, allowing us to test whether directed and random exploration increase with horizon. Consistent with previous research, we find that this is indeed the case (Figure \ref{fig:modelfree}C, D; STATS FOR CHANGE IN INFO BONUS AND DECISION NOISE).

	\added[id=bob]{Taken together our model-free and model-based analyses agree with previous findings showing increased behavioral variability and increased information seeking in the long horizon condition consistent with humans using random and directed exploration. However, for random exploration, this previous analysis cannot distinguish between deterministic and random sources of noise. For this we analyze the extent to which people's choices are consistent on the repeated games.}

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/RanDetNoise_modelfree.jpg}
		\caption{\added[id=bob]{ADD C AND D PANELS FOR MODEL-BASED RESULTS HERE}
			Replication of previous findings. Both  $p(\mbox{low mean})$ (A) and $p(\mbox{high info})$ (B) increase with horizon suggesting that people use both random and directed exploration in this task.  }
		\label{fig:modelfree}
	\end{center}
\end{figure}

@@

Next we asked whether participants' choices were consistent or inconsistent in the two repetitions of each game. The idea behind this measure is that purely deterministic noise should lead to consistent choices as the deterministic stimulus is identical both times. \deleted[id=siyu]{Conversely, if choice is not a deterministic function of the stimulus, participants' choices should be independent, and hence more inconsistent across the repetitions of the game.}\added[id=siyu]{Conversely, if choice is not entirely driven by a deterministic process and is also driven by random noise, participants' choices should be more inconsistent across the repetitions of the game. Moreover, if decision noise is purely random noise, meaning there is no unobserved deterministic process, we will show that we can actually predict the expected level of choice inconsistencies across repetitions of games by accounting for the known deterministic processes and assuming that the random noise process is independent in repetitions of the game.}

To quantify choice inconsistency we computed the frequency with which participants made different responses for pairs of repeated games (Figure \ref{fig:mf2}, Supplementary Figure S2). Using this measure we found that participants made inconsistent choices in both the unequal ([1 3]) and equal ([2 2]) information conditions, suggesting that not all of the noise was stimulus driven (t-test vs zero revealed that inconsistency was greater than zero for all horizon and uncertainty conditions\deleted[id=siyu]{For [1 3] condition, t(64) = 13.72, p $<$ 0.001 for horizon 1, t(64) = 16.71, p $<$ 0.001 for horizon 6; For [2 2] condition, t(64) = 9.55, p $<$ 0.001 for horizon 1, t(64) = 17.93, p $<$ 0.001 for horizon 6}). In addition, we found that choice inconsistency was higher in horizon 6 than in horizon 1 for both [1 3] and [2 2] condition (For [1 3] condition, t(64) = 5.41, p $<$ 0.001; for [2 2] condition, t(64) = 6.26, p $<$ 0.001), suggesting that at least some of the horizon dependent noise is not a deterministic function of the stimulus\added[id=siyu]{, but rather random noise}.

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/RanDetNoise_pinconsistent.jpg}
		\caption{Model-free analysis suggests that both deterministic and random noise contribute to the choice variability in random exploration. For both the [1 3] (A) and [2 2] (B) condition, people show greater choice inconsistency in horizon 6 than horizon 1. However, the extent to which their choices are inconsistent lies between what is predicted by purely deterministic and random noise, suggesting that both noise sources influence the decision.}
		\label{fig:mf2}
	\end{center}
\end{figure}

To gain more quantitative insight into these results, we computed theoretical values for the choice inconsistency for the purely deterministic and purely random noise cases.  For purely deterministic noise this computation is simple because people should make the exact same decisions each time in repeated games, meaning that $p(\mbox{inconsistent}) = 0$ in this case. For purely random noise, the two games should be treated independently, allowing us to compute the choice inconsistency in terms of the probability of choosing the low mean option, $p(\mbox{low mean})$, as
\begin{equation*}
	\begin{split}
		p(\mbox{consistent}) &= p(\mbox{low mean})^2 + p(\mbox{high mean})^2\\
		&= p(\mbox{low mean})^2 + (1-p(\mbox{low mean}))^2\\ 
		\mbox{hence},\quad p(\mbox{inconsistent}) &=  
		1 - p(\mbox{consistent}) = 
		2 p(\mbox{low mean})(1-p(\mbox{low mean}))
	\end{split}
\end{equation*}
\added[id=siyu]{Furthermore, to account for the fact that $p(\mbox{low mean})$ is a function of reward difference $\Delta R$ between the two bandits and the information condition $I$, we estimated the conditional probability:
	$$p(\mbox{inconsistent}|\Delta R, I) = 2 p(\mbox{low mean}|\Delta R, I)(1-p(\mbox{low mean}|\Delta R, I))$$
	Then based on the likelihood that each condition ($\Delta R$ vs $I$) occurs in the task $\rho(\Delta R, I)$, we have
	$$p(\mbox{inconsistent}) = \sum_{\Delta R, I}\rho(\Delta R, I)p(\mbox{inconsistent}|\Delta R, I)$$}

As shown in Figure \ref{fig:mf2}, people's behavior falls in between the pure deterministic noise prediction and the pure random noise prediction. Specifically, behavior is different from the pure random noise prediction in the both the [1 3] condition (t(64) = 4.83, p $<$ 0.001 for horizon 1, t(64) = 3.12 p $=$ 0.003 for horizon 6) and the [2 2] condition (t(64) = 3.92, p $<$ 0.001 for horizon 1, t(64) = 3.71, p $<$ 0.001 for horizon 6). Likewise, behavior is different from pure deterministic noise prediction in both the [1 3] condition (t(64) = 13.72, p $<$ 0.001 for horizon 1, t(64) = 16.71, p $<$ 0.001 for horizon 6) and the [2 2] condition (t(64) = 9.55, p $<$ 0.001 for horizon 1, t(64) = 17.93, p $<$ 0.001 for horizon 6). \added[id=bob]{As a negative control of our method for estimating $p(\mbox{inconsistent})$ for purely random noise, we simulated choices using a decision model that only includes random noise (Equation \ref{eq:origmodel}), and found that $p(\mbox{inconsistent})$ in this simulated data is not different from our pure random noise prediction in all horizon and uncertainty conditions (p $>$ 0.05, Supplementary Figure S3).} \replaced[id=siyu]{Together, our results suggest that both random noise and deterministic noise contribute to the choice variability in random exploration.}{This suggests that at least some of the `noise' is a deterministic function of the stimulus, }. However,  how each of these types of noise change with horizon is difficult to discern.

\subsection*{Model-based analysis \replaced[id=siyu]{provides a lower-bound estimate of deterministic noise and an upper-bound estimate of random noise}{shows deterministic noise changes with horizon}}

To more precisely quantify the contribution of deterministic noise \added[id=siyu]{and random noise}, we turned to model fitting. We modeled behavior on the first free choice of the Horizon Task using a version of the logistic choice model in \citep{wilson2014} (Equation \ref{eq:origmodel}) that was modified to differentiate \replaced[id=siyu]{between components of the noise that are deterministically driven by the stimulus (`deterministic noise') and components of the noise that are not deterministically driven by the stimulus (`random noise')}{deterministic noise from random noise}. In particular, we assume that in repeated games, the value of stimulus-driven deterministic noise is frozen whereas random noise is drawn independently both times.

\subsubsection*{Overview of model}
%As with our model-free analysis, the model-based analysis focuses only on the first free-choice trial since that is the only free choice when we have control over the experience participants have about two bandits. 
\added[id=bob]{To model participants' choices on the first free-choice trial, we use a modified version of Equation \ref{eq:origmodel}.
}
%assume that they make decisions by computing the difference in value $\Delta Q$ between the right and left options, choosing right when $\Delta Q > 0$ and left otherwise.  Specifically, we write
\begin{equation}
	\label{eq:newmodel}
	\Delta Q= \Delta R+A \Delta I+b+n_{det}+n_{ran}
\end{equation}
where, as before $\Delta R$, is the the difference in mean rewards shown on the forced trials, $\Delta I$, is the difference in information, $A$ is the information bonus, and $b$ is the spatial bias.
%\added[id=siyu]{of} information available for playing the two options on the first free-choice trial. %For simplicity, and because information is manipulated categorically in the Horizon Task, we define $\Delta I$ to be +1 if one reward is drawn from the right option and three are drawn from the left in the [1 3] condition, -1 if one from the left and three from the right, and in [2 2] condition, $\Delta I$ is 0. \added[id=siyu]{$n_{det}$ denotes the deterministic noise, which is identical on the repeat versions of each game; and $n_{ran}$ denotes random noise, which is uncorrelated between repeated plays and changes every game.} $n_{det}$ and $n_{ran}$ are assumed to come from logistic distributions with mean 0\added[id=siyu]{, and standard deviations $\sigma_{det}$ and $\sigma_{ran}$}. 
%The subject-and-condition-specific parameters are: the spatial bias, $b$, which determines the extent to which participants prefer the option on the right; the information bonus $A$, which controls the level of directed exploration; 
New in Equation \ref{eq:newmodel} are the terms $n_{det}$ and $n_{ran}$. $n_{det}$ denotes the deterministic noise, which is identical on the repeat versions of each game; and $n_{ran}$ denotes random noise, which is uncorrelated between repeated plays and changes every game. $n_{det}$ and $n_{ran}$ are assumed to come from logistic distributions with mean 0, and standard deviations $\sigma_{det}$ and $\sigma_{ran}$.

% $\sigma_{det}$ and $\sigma_{ran}$, which control the respective level of deterministic noise and random noise in random exploration.}

@@

\subsubsection*{\replaced[id=siyu]{Model validation}{Parameter recovery}\label{ch:appendix:bayesrecovery}}
To be sure that our fit parameter values were meaningful and to understand the limits of our model, we evaluated our model extensively using simulated data. \added[id=bob]{This allowed us to quantify whether deterministic and random noise can be identified under ideal conditions where the behavior is generated by the model with known parameters. Full details of this analysis are presented in the Supplementary Materials. }

	In this section we focus on our results for parameter recovery. In a parameter recover analysis, behavioral data is simulated by the model with known parameters and then this simulated behavioral data fit with the model to quantify the extent to which fit parameters match the input simulated parameters --- that is, whether the simulated parameters can be recovered.

	Parameter recover in this task was good for this model, with fit values for $\sigma_{ran}$ and $\sigma_{det}$ showing strong correlations with their simulated values (INSERT STATS HERE). However, while the relationship was near perfect for random noise, there was a systematic bias to underestimate the level of deterministic noise by about XXX (INSERT VALUE HERE). Despite this underestimation of deterministic noise in both horizon conditions, the difference in deterministic noise between horizons is much better captured (STATS). This is because the underestimation of deterministic noise is partially canceled out when the difference is taken between horizon conditions.

	INSERT A SIMPLE PARAMETER RECOVERY PLOT FOR SIGMARAN AND SIGMADET.

	%this simulated behavioral data is fit using the same fitting procedure used to model human choices.

@@

MOVE THE REST OF THIS SECTION TO THE SUPPLEMENTARY MATERIAL
Firstly, we checked if our fitted deterministic noise could indeed capture unobserved deterministic process that was not accounted for by the decision model. We test this by leaving out one known deterministic process from the decision model, and ask if our method could recover that known deterministic process as deterministic noise. In particular, we fit a reduced version of our model that only considers reward and ignores the influence of uncertainty condition on explore-exploit decisions.  
$$\Delta Q= \Delta R+n_{det}+n_{ran}$$
Here, $\Delta Q, \Delta R, n_{det}, n_{ran}$ represent the same variables as in the full model.	If deterministic noise in our model can indeed capture unobserved deterministic processes that's missed by the model, then we would expect to see a higher level of fitted deterministic noise in the reduced model compared to in the full model, whereas the level of random noise should remain unchanged. By comparing the	fitted posterior distributions over the group-level means of the deterministic and random noise parameters $\sigma_{det}$ and $\sigma_{ran}$, as expected, we observed an increase in deterministic noise and no change in random noise between the reduced and the full model (Figure \ref{fig:reducedmodel}). This suggests that our model is capable of detecting missing deterministic processes.

\begin{figure}[H]
\begin{center}
	\includegraphics[width=0.7\textwidth]{figures/RDBayes_reduced_model.jpg}
	\caption{\added[id=bob]{FONT SIZE ON LEGEND IS VERY SMALL!}Deterministic noise can recover known deterministic processes that's intentionally omitted by the model. In the reduced model where the deterministic effect of uncertainty condition is omitted from the model, deterministic noise is higher compared to the full model that accounts for the effect of uncertainty. Random noise remains unchanged between the two models.}
	\label{fig:reducedmodel}
\end{center}
\end{figure}

\added[id=siyu]{Secondly, we evaluated our hierarchical Bayesian analysis procedure using the `frequentist coverage analysis'. In the coverage analysis, we simulated choices with the fitted parameters from the Hierarchical Bayesian analysis, and then re-fit the simulated choices to see whether we can recover the parameters (Figure \ref{fig:coverage2}, Supplementary Figure S4). The simulation and re-fitting was repeated for 200 times. Then we counted out of the 200 repetitions how many times the true parameter that we simulate the choices from lies in the fitted 95\% confidence interval. If our model fitting is reliable, then the fitted confidence interval should cover the true parameter for more than 95\% of the simulations (this ratio will be referred to as the coverage rate). For random noise, the coverage rate is 100\% for both horizon 1, horizon 6, and the horizon difference. For deterministic noise, the coverage rate is 66\% for horizon 1 and 69\% for horizon 6. By comparing the posterior distributions of parameters that were used to generate simulations and the posterior distribution of recovered parameters, it is clear that our model systematically underestimates deterministic noise (Figure \ref{fig:coverage2}). Despite the underestimation of deterministic noise in both horizons, we could still reliably detect the horizon changes of deterministic noise (coverage rate is 97\%). This is because the underestimation of deterministic noise is partially canceled out when the difference is taken between horizons. For random noise, our model fitting procedure yields a faithful recovery. However, there is a conceptual limitation. Because random noise is modeled as non-stimulus-driven noise, it can include both true stochastic neural noise and possible deterministic noises which do not depend on the stimuli. Because of this, our random noise estimate provides an upper bound of true `random noise' induced by intrinsic stochastic processes in the brain.}

\begin{figure}[H]
\begin{center}
	\includegraphics[width=1\textwidth]{figures/RDBayes_parameterrecovery_hyperprior_v2.jpg}
	\caption{Parameter recovery over the posterior distribution of random and deterministic noise standard deviations $\sigma_{det}$ and $\sigma_{ran}$. Solid lines are true posterior used to simulate choices. Lighter color shades represent the re-fitted posterior to the simulated choices. Our model fitting procedure faithfully recovers the non-stimulus-driven random noise (A, B), but systematically underestimates deterministic noise in both horizons (D, E). The horizon differences in random noise is also faithfully recovered (C). The horizon differences in deterministic noise is also underestimated but not significant (F).}
	\label{fig:coverage2}
\end{center}
\end{figure}

\added[id=siyu]{Next, we tested the ability of our model fitting procedure to recovery parameters from simulated data at the subject level (Supplementary Figure S5 and S6). The correlations between the true vs fitted parameters are significant across participants for all parameters (p $<$ 0.001). The strength of correlation between simulated and fit values are strong for both deterministic noise (R $>$ 0.8) and random noise (R $>$ 0.9). Despite the strong inter-subject correlations, we again observed a systematic underestimation of $\sigma_{det}$ (Supplementary Figure S5 and S6).}

%Parameter recovery is good (in terms of) for all parameters apart from the bias in short horizon condition, which is likely due to this parameter being so close to zero (Figure \ref{fig:pararecover1} Panel C). The recovery for the noise parameters, $\sigma_{det}$ and $\sigma_{ran}$, is slightly better for horizon 1 than horizon 6.

\added[id=siyu]{Lastly, in addition to testing how our model performs in parameter ranges around the actual fitted parameters, we tested the limitations of our models in arbitrary combinations of random vs deterministic noises. All combinations of random and deterministic noises with $0 \le \sigma_{det} \le 10$ and $0 \le \sigma_{ran} \le 10$ were tested. In a special case, we evaluated how our model performs when there is only random noise or only deterministic noise (Figure \ref{fig:puredetran}). In the simulation with fully deterministic noise and 0 random noise, our model successfully recovered both random and deterministic noise (Figure \ref{fig:puredetran} C, D), however in the simulation with fully random noise and 0 deterministic noise, although our model successfully recovered random noise, some small proportion of deterministic noise was falsely detected when they should instead be 0 (Figure \ref{fig:puredetran} A, B). However, this phenomenon only exists when the true deterministic noise is 0, once the true deterministic noise is greater than 1, we don't observe this inflation of deterministic noise anymore (Supplementary Figures S7). Apart from this, our model did a fairly good job in recovering all combinations of random and deterministic noises (Supplementary Figures S7).}

%Overall, we see better recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games.

%This is because it requires more trials to recover larger values of $\sigma_{det}$ and $\sigma_{ran}$, so with the same number of choices it is harder to recover overall larger noise variances in horizon 6.

\begin{figure}[H]
\begin{center}
	\includegraphics[width=0.7\textwidth]{figures/RDBayes_parameterrecovery_grid_pureRanDet.jpg}
	\caption{Parameter recovery over the posterior of random noise standard deviation, $\sigma_{ran}$, and deterministic noise standard deviation, $\sigma_{det}$, for purely random noise (top row) and purely deterministic noise (bottom row) games. 
	}
	\label{fig:puredetran}
\end{center}
\end{figure} 
% Thus, our model provides a lower bound on deterministic noise and an upper bound on random noise.

\added[id=siyu]{Overall, we are able to detect both deterministic and random noises using our model to a satisfactory extent. Our model provides a lower bound for deterministic noise and an upper bound for random noise. In addition, We see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games. 
}
\subsubsection*{Model-based results}

>>>>

	\subsubsection*{Model-based analysis}	
	Another approach to understanding behavior in the Horizon Task is to use a computational model \citep{wilson2014}. In this case, we model  participants' choices on the first free-choice trial by assuming they make decisions by computing the difference in value (or utility) $\Delta Q$ between the right and left options, choosing right when $\Delta Q > 0$ and left otherwise.  Specifically, we write
	\begin{equation}
		\label{eq:origmodel}
		\Delta Q= \Delta R+A \Delta I+b+n
	\end{equation}
	where, the experimentally controlled variables are $\Delta R=R_{right}-R_{left}$, the difference between the mean of rewards shown on the forced-choice trials, and $\Delta I$, the difference in information available for playing the two options on the first free-choice trial. For simplicity, and because information is manipulated categorically in the Horizon Task, we define $\Delta I$ to be +1 if one reward is drawn from the right option and three are drawn from the left in the [1 3] condition, -1 if one from the left and three from the right in the [1 3] condition, and in [2 2] condition, $\Delta I$ is 0.

@@

	The free parameters of this model are: the information bonus $A$, which controls the level of directed exploration; the noise standard deviation, $\sigma$, which controls the level of random exploration, and the spatial bias, $b$, which determines the extent to which participants prefer the option on the right. These free parameters are fit separately for each participant in each horizon and information condition, allowing us to test whether directed and random exploration increase with horizon. Consistent with previous research, we find that this is indeed the case (Figure \ref{fig:modelfree}C; t(64) = 5.35, p $<$ 0.001. Figure \ref{fig:modelfree}D; t(64) = 3.54, p $<$ 0.001).

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/RanDetNoise_modelfree.jpg}
			\caption{Replication of previous findings that people use both random and directed exploration in this task. (A) model-free measure of behavioral variability, $p(\mbox{low mean})$, increases with horizon. (B) model-free measure of information seeking, $p(\mbox{high info})$, increases with horizon. (C) model-based measure of behavioral variability, decision noise $\sigma$, increases with horizon. (B) model-based measure of information seeking, information bonus $A$, increases with horizon.}
			\label{fig:modelfree}
		\end{center}
	\end{figure}

	Taken together, our model-free and model-based analyses agree with previous findings showing increased behavioral variability and increased information seeking in the long horizon condition, consistent with humans using random and directed exploration (Figure \ref{fig:modelfree}, Supplementary Figure S1). However, for random exploration, this previous analysis cannot distinguish between deterministic and random sources of noise. For this we analyze the extent to which people's choices are consistent on the repeated games.

@@

@@

Next we asked whether participants' choices were consistent or inconsistent in the two repetitions of each game. The idea behind this measure is that purely deterministic noise should lead to consistent choices as the deterministic stimulus is identical both times. Conversely, if choice is not entirely driven by a deterministic process and is also driven by random noise, participants' choices should be more inconsistent across the repetitions of the game. Moreover, if decision noise is purely random noise, meaning there is no unobserved deterministic process, we will show that we can actually predict the expected level of choice inconsistencies across repetitions of games by accounting for the known deterministic processes and assuming that the random noise process is independent in repetitions of the game.

To quantify choice inconsistency we computed the frequency with which participants made different responses for pairs of repeated games (Figure \ref{fig:mf2}, Supplementary Figure S2). Using this measure we found that participants made inconsistent choices in both the unequal ([1 3]) and equal ([2 2]) information conditions ($p(\mbox{inconsistent}) > 0 $), suggesting that not all of the noise was stimulus driven. In addition, we found that choice inconsistency was higher in horizon 6 than in horizon 1 for both [1 3] and [2 2] condition (For [1 3] condition, t(64) = 5.41, p $<$ 0.001; for [2 2] condition, t(64) = 6.26, p $<$ 0.001), suggesting that at least some of the horizon dependent noise is not a deterministic function of the stimulus, but rather random noise.

% (t-test vs zero revealed that inconsistency was greater than zero for all horizon and uncertainty conditions)

%\deleted[id=siyu]{For [1 3] condition, t(64) = 13.72, p $<$ 0.001 for horizon 1, t(64) = 16.71, p $<$ 0.001 for horizon 6; For [2 2] condition, t(64) = 9.55, p $<$ 0.001 for horizon 1, t(64) = 17.93, p $<$ 0.001 for horizon 6}
\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/RanDetNoise_pinconsistent.jpg}
		\caption{Model-free analysis suggests that both deterministic and random noise contribute to the choice variability in random exploration. For both the [1 3] (A) and [2 2] (B) condition, people show greater choice inconsistency in horizon 6 than horizon 1. However, the extent to which their choices are inconsistent lies between what is predicted by purely deterministic and random noise, suggesting that both noise sources influence the decision.}
		\label{fig:mf2}
	\end{center}
\end{figure}

To gain more quantitative insight into these results, we computed theoretical values for the choice inconsistency for the purely deterministic and purely random noise cases.  For purely deterministic noise this computation is simple because people should make the exact same decisions each time in repeated games, meaning that $p(\mbox{inconsistent}) = 0$ in this case. For purely random noise, the two games should be treated independently, allowing us to compute the choice inconsistency in terms of the probability of choosing the low mean option, $p(\mbox{low mean})$, as
\begin{equation*}
	\begin{split}
		p(\mbox{consistent}) &= p(\mbox{low mean})^2 + p(\mbox{high mean})^2\\
		&= p(\mbox{low mean})^2 + (1-p(\mbox{low mean}))^2\\ 
		\mbox{hence},\quad p(\mbox{inconsistent}) &=  
		1 - p(\mbox{consistent}) = 
		2 p(\mbox{low mean})(1-p(\mbox{low mean}))
	\end{split}
\end{equation*}

Furthermore, to account for the fact that $p(\mbox{low mean})$ is a function of reward difference $\Delta R$ between the two bandits and the information condition $I$, we estimated the conditional probability:
	$$p(\mbox{inconsistent}|\Delta R, I) = 2 p(\mbox{low mean}|\Delta R, I)(1-p(\mbox{low mean}|\Delta R, I))$$
	Then based on the likelihood that each condition ($\Delta R$ and $I$) occurs in the task $\rho(\Delta R, I)$, we have
	$$p(\mbox{inconsistent}) = \sum_{\Delta R, I}\rho(\Delta R, I)p(\mbox{inconsistent}|\Delta R, I)$$

As shown in Figure \ref{fig:mf2}, people's behavior falls in between the pure deterministic noise prediction and the pure random noise prediction. Specifically, behavior is different from the pure random noise prediction in the both the [1 3] condition (t(64) = 4.83, p $<$ 0.001 for horizon 1, t(64) = 3.12 p $=$ 0.003 for horizon 6) and the [2 2] condition (t(64) = 3.92, p $<$ 0.001 for horizon 1, t(64) = 3.71, p $<$ 0.001 for horizon 6). Likewise, behavior is different from pure deterministic noise prediction in both the [1 3] condition (t(64) = 13.72, p $<$ 0.001 for horizon 1, t(64) = 16.71, p $<$ 0.001 for horizon 6) and the [2 2] condition (t(64) = 9.55, p $<$ 0.001 for horizon 1, t(64) = 17.93, p $<$ 0.001 for horizon 6). As a negative control of our method for estimating $p(\mbox{inconsistent})$ for purely random noise, we simulated choices using a decision model that only includes random noise (Equation \ref{eq:newmodel}), and found that $p(\mbox{inconsistent})$ in this simulated data is not different from our pure random noise prediction in all horizon and uncertainty conditions (p $>$ 0.05, Supplementary Figure S3). Together, our results suggest that both random noise and deterministic noise contribute to the choice variability in random exploration. However, the relative contribution from each of these types of noise, as well as how each type of noise changes with horizon, are difficult to discern.

\subsection*{Model-based analysis provides a lower-bound estimate of deterministic noise and an upper-bound estimate of random noise}

To more precisely quantify the contribution of deterministic noise and random noise, we turned to model fitting. We modeled behavior on the first free choice of the Horizon Task using a version of the logistic choice model (Equation \ref{eq:origmodel}) that was modified to differentiate between components of the noise that are deterministically driven by the stimulus (`deterministic noise') and components of the noise that are not deterministically driven by the stimulus (`random noise'). In particular, we assume that in repeated games, the value of stimulus-driven deterministic noise is frozen whereas random noise is drawn independently both times.

\subsubsection*{Overview of model}
%As with our model-free analysis, the model-based analysis focuses only on the first free-choice trial since that is the only free choice when we have control over the experience participants have about two bandits. 
To model participants' choices on the first free-choice trial, we use a modified version of Equation \ref{eq:origmodel}.
%assume that they make decisions by computing the difference in value $\Delta Q$ between the right and left options, choosing right when $\Delta Q > 0$ and left otherwise.  Specifically, we write
\begin{equation}
	\label{eq:newmodel}
	\Delta Q= \Delta R+A \Delta I+b+n_{det}+n_{ran}
\end{equation}
where, as before $\Delta R$, is the the difference in mean rewards shown on the forced-choice trials, $\Delta I$, is the difference in information, $A$ is the information bonus, and $b$ is the spatial bias. New in Equation \ref{eq:newmodel} are the terms $n_{det}$ and $n_{ran}$. $n_{det}$ denotes the deterministic noise, which is identical on the repeat versions of each game; and $n_{ran}$ denotes random noise, which is uncorrelated between repeated plays and changes every game. $n_{det}$ and $n_{ran}$ are assumed to come from logistic distributions with mean 0, and standard deviations $\sigma_{det}$ and $\sigma_{ran}$.

%The subject-and-condition-specific parameters are: the spatial bias, $b$, which determines the extent to which participants prefer the option on the right; the information bonus $A$, which controls the level of directed exploration;

@@

\subsubsection*{Model validation \label{ch:appendix:bayesrecovery}}
To be sure that our fit parameter values were meaningful and to understand the limits of our model, we evaluated our model extensively using simulated data. This allowed us to quantify whether deterministic and random noise can be identified under ideal conditions where the behavior is generated by the model with known parameters. Full details of this analysis are presented in the Supplementary Materials.

In this section we focus on our results for parameter recovery \citep{Wilson2019}. In a parameter recover analysis, behavioral data is simulated by the model with known parameters and then this simulated behavioral data is fit with the model to quantify the extent to which fit parameters match the input simulated parameters --- that is, whether the simulated parameters can be recovered.

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/RDBayes_parameterrecovery_subject_examplesession_randet.jpg}
		\caption{Parameter recovery analysis for random (A,B) and deterministic noise (C,D) in the two horizons.}
		\label{fig:paramrecover_main}
	\end{center}
\end{figure}

Parameter recover in this task was good for this model, with fit values for $\sigma_{ran}$ and $\sigma_{det}$ showing strong correlations with their simulated values in both horizon (H) conditions (For $\sigma_{ran}$, $R = 0.91 (\mbox{H} = 1) \mbox{ and } 0.84 (\mbox{H} = 6), p < 0.001$, For $\sigma_{det}$, $R = 0.60 (\mbox{H} = 1) \mbox{ and } 0.59 (\mbox{H} = 6), p < 0.001$). However, while the relationship was near perfect for random noise ($\frac{\mbox{\small Recovered } \sigma_{ran}}{\mbox{\small Simulated } \sigma_{ran}} = 1.01$), there was a systematic bias to underestimate the level of deterministic noise by about 32\% ($\frac{\mbox{\small Recovered } \sigma_{det}}{\mbox{\small Simulated } \sigma_{det}} = 0.68$). Despite this underestimation of deterministic noise in both horizon conditions, the difference in deterministic noise between horizons is much better captured (Supplementary Figure XXX). This is because the underestimation of deterministic noise is partially canceled out when the difference is taken between horizon conditions.
	% $\frac{\mbox{\small Recovered } \Delta \sigma_{det}}{\mbox{\small Simulated } \Delta \sigma_{det}} = 0.74$

@@

\subsubsection*{Model-based results}
